Title

### 2022-04-27 thesis/intro/intro.tex
carlos revision 1

\section{Theoretical Background} \label{sec:theoretical_background}

A model is a formal representation of a natural system. The representation of computer system models includes equations, graphical models, rules, decision trees, representative collections of examples, and neural networks. The choice of representation affects the model's accuracy, as well as its interpretability by people~\cite{Hypothesis2012EncyclopediaLearning, Roy2019ForecastingNetwork, Zhu2019PredictingLearning}. Accurate energy and power consumption models are essential for many energy efficiency schemes employed in computing equipment \cite{Rivoire2007ModelsOptimizations}, and they can have multiple uses, including the design, forecasting, and optimization of data center systems. This study focuses on analytical models that could aid energy optimization and analyses of crucial factors in the total energy draw.

The desirable properties of a full-system model of energy consumption include accuracy, speed, generality and portability, inexpensiveness, and simplicity \cite{Rivoire2008AModels}. However, modeling an HPC system's exact energy consumption behavior is not straightforward, either at the whole-system level or at the level of individual components. Data centers' energy consumption patterns depend on multiple factors, such as hardware specifications, workload, cooling requirements, or the type of the applications. Some of these factors cannot be measured easily. Furthermore, it is impractical to perform detailed measurements of the energy consumption of lower-level components without additional overhead.

Several proposed models have already been classified concerning their input parameters, as shown by Dayarathna et al.~\cite{Dayarathna2016DataSurvey}, who analyzed more than 200 models according to their characteristics and limitations and classified them into categories where the model is more suited to its objectives:
\begin{itemize}
	%\item Temperature
	\item System utilization or workload
	\item Frequency
	\item Other system states, such as cache miss, branch prediction, number of instructions executed, and more
	\label{tab:input_type}
\end{itemize}

Often, energy models are described as a combination of two main parts, the power model of the system and the performance model of the application. This is because the concept of energy (E) is the total amount of work performed by a system over a period of time (T), while power (P) is the rate at which the system performs the work. The relation between these three amounts can be expressed as:
\begin{equation}
	E = \int_{0}^{T}P(t)dt.
	\label{eq:energy_definition_cont}
\end{equation}

\subsection{Power Models}

The modeling of system parameters is becoming popular nowadays with the advantage of performance counters provided by the CPU or the operating system. These counters can measure micro-architectural events, such as instructions executed, cache hits, miss-predicted branches, and more; thus, providing a base for many different estimations of power usage. This makes this type of model very suitable for power estimation because it can use information about several internal states of the computer.

Frequency-based models are the most common kind of model. They serve as a base for many power models~\cite{Sarwar1997CmosCalculation, Butzen2007LeakageGates, Usman2013ANoC}. These models utilize the fact that every digital circuit (including modern processors) is composed of transistors. Thus, modeling one transistor's interaction and scaling this to the chip can give a reasonable estimate of the entire system's energy. One of the most common frequency-based %please confirm that the intended meaning has been retained.
model approximations is defined as follows: 
\begin{equation}
	P = \alpha+\beta f^3,
	\label{eq:power_simplified}
\end{equation}
where $\alpha$ and $\beta$ are model parameters, and $f$ is the operating frequency (details of this equation are covered in (\cref{chapter:models}). This type of model is suitable for optimization problems since these are a function of the operating frequency, which can be easily controlled.

% We confirm that the meaning has been retained

\subsection{Performance Models}
The most common way to model the application performance is using the workload. The workload is an abstract representation of the amount of work done for a given time and speed. The workload ($W$) can be defined in many different ways. One common way, used in many studies, such as Paolillo et al. ~\cite{Paolillo2018OptimisationParallelism}, Francis et al. \cite{ Group2012HandbookSahni}, and Kim et al. \cite{Kim2015RacingHeuristics}, is the following:
\begin{equation}
	W = \int_{0}^{\tau}s(t)dt = s\tau,
	\label{eq:workload_definition}
\end{equation}
where $\tau$ is total active time, and $s$ is the execution speed in instructions/second.

Utilization models \cite{Fu2018RaceMinimization, Group2012HandbookSahni} are also found in the literature, defined as the ratio between the time that the system is active and the total time (idle and active). These models are present in many DVFS algorithms present in Linux. They can be viewed as a good alternative to the workload since it is impossible to measure workload in real-time.  \cref{eq:utilization_definition} defines workload in terms of CPU utilization ($u$):
\begin{equation}
	u = \frac{\tau}{T} = \frac{W/s}{T},
	\label{eq:utilization_definition}
\end{equation}
where $T$ is the total execution time (idle and active), and $\tau$ is the active time, meaning when the processor was executing instructions.
Models based on CPU utilization are the basis for DVFS algorithms. Even though this is not a controllable parameter, it is straightforward to measure system utilization with almost no overhead, and it is also very portable in terms of operating systems and architectures.

\section{Related work} \label{sec:related_work}

Merkel et al.~\cite{Merkel2006BalancingSystems} developed an energy model for processors based on events. Their model assumes a fixed energy consumption $\alpha_i$ for each activity, and by counting the number of occurrences $c_i$ of every activity, they estimate the total energy as:
\begin{equation}
	E = \sum_{i=1}^{n}\alpha_ic_i.
	\label{equation:rw_merkel}
\end{equation}

Another event-based model, introduced by Roy et al. \cite{Roy2013AnAlgorithms}, described the computational energy consumed by a CPU for an algorithm $A$ as the Equation (\ref{equation:rw_event_based})%\cref{equation:rw_event_based}:
\begin{equation}
	E(A) = P_{clk}T(A) + P_wW(A),
	\label{equation:rw_event_based}
\end{equation}
where $P_{clk}$ is a processor clock leakage power, $T(A)$ is the total execution time, $W(A)$ is the total time taken by  non-I/O operations, and $P_w$ is used to capture the power consumption per operation performed by the CPU. $T(A)$ and $W(A)$ are estimated using performance features.

Models based on events present some drawbacks, they are highly dependent on the operating system and its architecture, making them problematic to port for other platforms. There are also limitations regarding the number of simultaneous events that can coexist without adding a non-negligible overhead. Additionally, there are cases where events need multiplexing, for example, when using more hardware events than %please confirm that the intended meaning has been retained.
the CPU can provide. There are also some well know problems regarding the precision of some events, as shown in many studies~\cite{Weaver2008CanTrusted, Weaver2013Non-determinismImplementations, Das2019SoK:Security, McGuire2009AnalysisKernel, Ramos2019AnCounters, Silva-de-Souza2020ContainergyAWorkloads}. Some events that should be exact and deterministic (such as the number of executed instructions) show run-to-run variations and over-count on various architectures, even when running in strictly controlled environments. Because of that, our proposed model is not dependent on events and, therefore, not vulnerable to those drawbacks.
% We confirm that the meaning is the same

An instruction-level energy model was also proposed in \cite{Shao2013EnergyProcessor} by Yakun et. al. Where they proposed an energy per instruction (EPI) characterization made on Xeon Phi. Their model is expressed as:
\begin{equation}
	E(f) = \frac{(p_1 - p_0)(c_1 - c_0)/f}{N}, 
	\label{eq:rw_instruction_level}
\end{equation}
where $N$ is the total number of dynamic instructions, $p_0$ is the initial idle power, $p_1$ is the average dynamic power, and ($c_1$ $-$ $c_0$) refers to the cumulative number of cycles the micro-benchmark performs. This model is suitable for estimating the energy after the application finishes executing when it is possible to count the total cycles. However, it is challenging to use for optimization or forecasting since it does not have an application model to predict the cycles. Our model integrates the behavior of the application, taking into account the execution time.

Lewis et al.~\cite{Lewis2008Run-timeSystems} described the overall system energy consumption using the following equation:
\begin{equation}
	E = A_0(E_{proc} + E_{mem}) + A_1E_{em} + A_2E_{board} + A_3E_{hdd},
	\label{eq:lewvis}
\end{equation}
where, $A_0$, $A_1$, $A_2$, and $A_3$ are unknown constants that are calculated via linear regression analysis and those remain constant for a specific server architecture. This model, as the previous one, relies on knowledge of energy spent on each component, being  a suitable option for estimation after the application has already run, but not for optimization of the run itself, which is the aim of our model.

In another energy consumption model  based on system utilization, Mills et al. \cite{Mills2014EnergySystems} modeled the energy consumed by a compute node with CPU (single) executing at speed $\sigma$ as Equation (\ref{eq:mills})%\cref{eq:mills},
\begin{equation}
	E(\sigma,[t_1,t_2]) = \int_{t_1}^{t_2} \sigma^3 + \rho \sigma_max^3 dt,
	\label{eq:mills}
\end{equation}
where $\rho$ stands for the overhead power consumed regardless of the processor speed, $t_1$ and $t_2$ are the application's initial and final execution times. The overhead includes power consumption by all other system components, such as memory, network, and more. For this reason, although the authors mentioned the energy consumption of a socket, their power model is generalized to the entire server. This model lacks a closed-form, i.e., it depends on the definition of $\alpha(t)$ to be complete. Our model has a closed-form which facilitates analyses.

\subsection{The contribution of this work}

@@

%The main contributions of the proposed model are:
%
%\begin{itemize}
%	\item Simple model: faster to fit and compute, good for DVFS and DPM optimization.
%	\item Parameters with logical meaning: helps to understand the contribution of each specific term.
%	\item Analytical analysis: several analyses can be derived from the equation.
%	\item Controllable variables: the equation is in the function of parameters that we can control directly.
%\end{itemize}

>>>>

\section{Objectives}

\section{Contributions}

@@

The main contributions of the proposed model are:

% Contribuições por capitulo...
\begin{itemize}
	\item Simple model: faster to fit and compute, good for DVFS and DPM optimization.
	\item Parameters with logical meaning: helps to understand the contribution of each specific term.
	\item Analytical analysis: several analyses can be derived from the equation.
	\item Controllable variables: the equation is in the function of parameters that we can control directly.
\end{itemize}

\section{Organization}

### 2022-04-27 thesis/phases/phases.tex
add journals and restructured
carlos revision 1

\section{The impact of best-choice execution time slots partition of HPC applications on energy-saving strategies}

@@

\subsection{Introduction}
Talk about the importance of energy efficiency.

\subsection{Related work}
Some related work

\subsection{Methodology}

\subsubsection{Optimal energy configuration}
Splitting a program into phases, or time intervals with a selected set of execution parameters, has an impact on energy consumption. To find the ideal phase division, we need to compute the optimal energy configuration for a given number and duration of phases. For that, many different approaches could be applied, each having different trade-offs.

@@

\subsubsection{Data Collection}
Data, power samples, can be gathered by running the application in several possible context configurations. In a system with 32 cores that can run at 13 frequencies, we can have 416 (32*13) possible configurations. Power samples can be taken from dedicated sensors in the IPMI system [REF]. To avoid overload, a sample rate of 0.5 seconds is more than sufficient. Additionally, the behavior of an application can be affected by the workload, so it is also interesting to run applications with different inputs and assess their impact on the collected data [REF victor alex paper].

@@

\subsubsection{The simulator}
To analyse the impact of configurations and identify optimal time-intervals, phases, we propose an integrator, which computes the energy on an interval for a given configuration, and a reference, a brute force test of all possible combinations. The integrator has order zero, and computers the energy between two timestamps, as shown in the figure \ref{fig:zero_order}, where the red and yellow parts area are the energy value.

\begin{figure}[h]
    \centering
    \includegraphics[width=\columnwidth]{phases/figures/zero_order.png}
    \caption{Power vs time intervals for a given application. The energy associated area is colored.}
    \label{fig:zero_order}
\end{figure}

@@

\subsubsection{Optimizing phases}
The main goal of this section is to identify the optimal number and duration of time intervals. Indeed, several short phases can have an impact on switching overhead as well as few long phases on the overall energy.

@@

\subsection{Results}

@@

\begin{figure}[h]
\includegraphics[width=\columnwidth]{fingerprint/figures/energy_per_phase.png}
    \caption{Relative energy vs number of phases using 13 applications from PARSEC 3.0, HPC, and Openmc benchmarks. Energy compared to the single-phase optimal configuration.}
    \label{fig:relative_energy}
\end{figure}

@@

The images below show the phases for all applications:
\begin{figure}[h]
\includegraphics[width=\columnwidth]{phases/figures/phase_division_35.png}
\end{figure}
In most applications, we can see a pattern of 3 phases wich for this applications are generally, and setup phase where it loads data from the disk, a computation phase where the data is processed, and a finalization phase where data is written back to the disk or display to the user.

\begin{figure}[h]
\includegraphics[width=\columnwidth]{phases/figures/phase_division_3.png}
\end{figure}
Plotting for 3 phases, we can observe that better.

\subsection{Conclusion and future work}
The main conclusion of this work was that in HPC enviroment we don't need a lot of phases to optimize the energy consumption given that for one phase we can compute the optimal energy.

>>>>

\section{Model with phases}

@@

\section{Introduction}
Talk about the importance of energy efficiency.

\section{Related work}
Some related work

\section{Methodology}

\section{Optimal energy configuration}
Splitting a program into phases, or time intervals with a selected set of execution parameters, has an impact on energy consumption. To find the ideal phase division, we need to compute the optimal energy configuration for a given number and duration of phases. For that, many different approaches could be applied, each having different trade-offs.

@@

\section{Data Collection}
Data, power samples, can be gathered by running the application in several possible context configurations. In a system with 32 cores that can run at 13 frequencies, we can have 416 (32*13) possible configurations. Power samples can be taken from dedicated sensors in the IPMI system [REF]. To avoid overload, a sample rate of 0.5 seconds is more than sufficient. Additionally, the behavior of an application can be affected by the workload, so it is also interesting to run applications with different inputs and assess their impact on the collected data [REF victor alex paper].

@@

\section{The simulator}
To analyse the impact of configurations and identify optimal time-intervals, phases, we propose an integrator, which computes the energy on an interval for a given configuration, and a reference, a brute force test of all possible combinations. The integrator has order zero, and computers the energy between two timestamps, as shown in the figure \ref{fig:zero_order}, where the red and yellow parts area are the energy value.

\begin{figure}[H]
    \centering
    \includegraphics[width=\columnwidth]{phases/figures/zero_order.png}
    \caption{Power vs time intervals for a given application. The energy associated area is colored.}
    \label{fig:zero_order}
\end{figure}

@@

\section{Optimizing phases}
The main goal of this section is to identify the optimal number and duration of time intervals. Indeed, several short phases can have an impact on switching overhead as well as few long phases on the overall energy.

@@

\section{Results}

@@

\begin{figure}[H]
\includegraphics[width=\columnwidth]{fingerprint/figures/energy_per_phase.png}
    \caption{Relative energy vs number of phases using 13 applications from PARSEC 3.0, HPC, and Openmc benchmarks. Energy compared to the single-phase optimal configuration.}
    \label{fig:relative_energy}
\end{figure}

@@

The images below show the phases for all applications:
\begin{figure}[H]
\includegraphics[width=\columnwidth]{phases/figures/phase_division_35.png}
\end{figure}
In most applications, we can see a pattern of 3 phases wich for this applications are generally, and setup phase where it loads data from the disk, a computation phase where the data is processed, and a finalization phase where data is written back to the disk or display to the user.

\begin{figure}[H]
\includegraphics[width=\columnwidth]{phases/figures/phase_division_3.png}
\end{figure}
Plotting for 3 phases, we can observe that better.

\section{Conclusion and future work}
The main conclusion of this work was that in HPC enviroment we don't need a lot of phases to optimize the energy consumption given that for one phase we can compute the optimal energy.
